Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-01-03 Manuscript/manuscript.Rmd
!!!separated paragraphs in intro

A study published by Mourad et al. used multiple logistic regression with LASSO regularization to predict TAD boundaries [@mourad2016computational]. They considered binning the genome at both 50 base and 1kb intervals and flanking TAD boundaries by 1kb and 20kb regions. The outcome Y was determined by whether or not a genomic bin overlapped with a flanked boundary. The feature space was obtained by considering the percent overlap between a bin and coordinates defining ChIP-seq peak data. Models were then evaluated using AUROC. However, a pervasive issue here becomes the introduction of highly imbalanced classes. Due to the sparcity of TADs throughout the genome, most bins will not overlap with a flanked TAD boundary. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@estabrooks2004multiple]. Similarly, careful consideration must also be taken when deciding which performance metric(s) to present when evaluating predictive models, especially in the case of imbalanced data. The popularity surrounding the use of ROC curves is that they do not depend on a threshold. However, when the prevalence of an event is low--that is, a genomic bin overlapping with a TAD boundary region--mean probabilities are biased towards the majority class. Likewise, metrics like accuracy and sensitivity and specificity do not incorporate the entirety of the confusion matrix created when validating predictive models on a test data set. Therefore, each of these metrics are likely to be inflated when evaluating model performance. [??? Make paragraphs contain one meaning. Split this paragraph into two, one describes the class imbalance problem, another is Mourad paper and how they didn't address it.]

@@

Furthermore, there are other aspects of TAD boundary modelling that current methods have failed to consider. Most notably is the concept of distance. As a feature engineering step, the use of distances (in base pairs) from genomic bins to regions of functional genomic elements has largely been ignored. However, these distance characteristics have the potential to offer a more accurate spatial representation of which genomic features are associated with TAD boundaries. Additionally, model performance is heavily contingent upon the resolution of the contact matrix used to identify the location of TADs throughout the genome. It is often unclear from authors of previous studies what resolution contact matrices were when retrieving TAD boundaries, and how other resolutions would affect their results. A larger resolution contact matrix will result in the identification of larger-sized TADs, and subsequently, less TAD boundary points. This will contribute to smaller sample sizes and could effect predictive performances.

@@

@@

>>>>

Current methods rely on the use of classification algorithms to identify which functional genomic elements are best able to discriminate between regions that include TAD boundaries and those that do not. Most methods differ on the techniques used to identify genomic regions that include TAD boundaries, as well as the models and tools used to evaluate the predictive capability of said models. In doing so, many methods ignore key characteristics of domain data that can impair and bias results. Most notably is the introduction of heavily imbalanced classes. Due to the sparcity of TADs throughout the genome, most defined regions will not contain a TAD boundary. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@estabrooks2004multiple]. Similarly, careful consideration must also be taken when deciding which performance metric(s) to present when evaluating predictive models, especially in the case of imbalanced data. The popularity surrounding the use of ROC curves is that they do not depend on a threshold. However, when the prevalence of an event is low--that is, a defined genomic region containing a TAD boundary--mean probabilities are biased towards the majority class. Likewise, metrics like accuracy and sensitivity and specificity do not incorporate the entirety of the confusion matrix created when validating predictive models on a test data set. Therefore, each of these metrics are likely to be inflated when evaluating model performance.

A study published by Mourad et al. used multiple logistic regression with LASSO regularization to predict TAD boundaries [@mourad2016computational]. They considered binning the genome at both 50 base and 1kb intervals and flanking TAD boundaries by 1kb and 20kb regions. The outcome Y was determined by whether or not a genomic bin overlapped with a flanked boundary. The feature space was obtained by considering the percent overlap between a bin and coordinates defining ChIP-seq peak data. Models were then evaluated using AUROC. Here, the authors ignored the effect of imbalanced classes as well as the use of potentially biased performance metrics.

[!!! Separated paragraphs]

@@

Furthermore, there are other aspects of TAD boundary modelling that current methods have failed to consider. Most notably is the concept of distance. As a feature engineering step, the use of distances (in base pairs) from genomic bins to regions of functional genomic elements has largely been ignored. However, these distance characteristics have the potential to offer a more accurate spatial representation of which genomic features are associated with TAD boundaries. Additionally, model performance is heavily contingent upon the resolution of the contact matrix used to identify the location of TADs throughout the genome. It is often unclear from authors of previous studies what resolution contact matrices were used when retrieving TAD boundaries, and how other resolutions would affect their results. A larger resolution contact matrix will result in the identification of larger-sized TADs, and subsequently, less TAD boundary points. This will contribute to smaller sample sizes and could effect predictive performances.

@@

## Figure 2: Predictor Types

Diagram of the 3 predictor types considered when assessing the relationship between TAD boundaries and functional genomic elements. Each predictor type was used as the feature space in downstream analyses for predicting which functional genomic elements were associated with the formation of TAD boundaries. Featured above are bin-specific examples of the construction of each predictor type. (Left) The overlap count (OC) predictors were calculated by considering the total number of elemental regions that overlapped with each genomic bin. (Middle) The overlap percent (OP) predictors were calculated by dividing the sum of all feature widths within each specific bin and dividing by the total bin width (either 5, 25, 50, or 100 kilobases given resolution of boundary data). (Right) The distance predictors were calculated by measuring the distance (in base pairs) from the center of each genomic bin to the center of the nearest elemental region of interest. The two green segments directly below the rightmost enlarged figure represent the center of the overlapping regions defining the respective functional genomic element.

@@

## Table 1: List of Genomic Annoations.

The full list of genomic annotations used in downstream analyses. These annotations were the functional genomic elements used to predict the formation of TAD boundaries. They were obtained via ChIP-Seq experiments, mapped to human genome assembly hg19, and made available by the ENCODE Consortium. Each element is denoted by its genomic class, and which cell type it is applied to. The elements that are invariant to a specific cell type are denoted by "Invariant". Additionally, a breif description and specific paths for downloads are also provided.
